|

### 2022-03-30 01-Chp-Background.Rmd
Range of corrections

@@

Dementia also has a substantial economic impact. In 2015, the estimated total cost of dementia in England was £24.2 billion. Health care costs alone were £3.8 billion, with the remainder being divided between unpaid care and social care costs.[@wittenberg2019] Globally, the cost of dementia care is expected to rise to $1tr by 2030.[@prince2014] As such, the urgent need to reduce the burden of dementia, both at the personal and system level, is clear.

@@

Given the substantial burden that dementia represents and the absence of any curative therapies, as detailed in the above sections, the assessment of easily modifiable targets for their utility in the prevention of dementia should be prioritized.[@winblad2016] To date, a substantial amount of research has been produced examining putative risk factors for dementia.[@feingold2000;@peters2019; @anstey2019]

<!-- TODO Need some details on risk factors here - and consider spacing differentl -->

The benefits of a prevention-based approach based on addressing these risk factors are well-studied. Reducing the prevalence of the seven most important risk factors for dementia (obesity, hypertension,[@hughes2020] diabetes, smoking, physical inactivity, and low educational attainment) by 10-20% per decade is estimated to result in a reduction in dementia prevalence of 8-15% by 2050.[@norton2014potential]

In this context, lipid levels represent a promising target for preventative treatment, due to the ready availability of lipid-modifying treatments which could be repurposed.[@pushpakom2019] Determining whether variations in lipid levels are causative for dementia may prove critical in reducing the future burden of the condition.

@@

Both trials were also limited by the relatively short follow-up period examined. This limited follow-up was to be expected because the primary aim of both trials was to assess the effect of statins on short-term coronary-related outcomes.[@heartprotectionstudycollaborativegroup2002; @trompet2010] The PROSPER trial had a mean follow-up of 3.2 years, while the MRC/BHF Protection Study estimated risk at 5 years of follow-up. Given the long lag time between non-symptomatic onset of dementia and clinical presentation, it is likely that these durations are insufficient to fully capture the onset of dementia. [Further]{.correction}, as they included only patients at high vascular risk, their generalisability to other settings is limited.[@mcguinness2016]

[Finally, the ability of studies of statin use to assess a causal link between lipid levels and dementia is limited by the pleiotropic effect of the intervention, as statins may effect dementia risk via some non-lipid-lowering pathway. For example, statins influence the levels of C-reactive protein,]{.correction}[@albert2001] [an independent risk factor for dementia.]{.correction}[@lewis2021; @schmidt2002]

@@

Recent MR studies indicated that genetically determined low levels of LDL-c may cause a reduction in AD risk.[@larsson2017; @ostergaard2015] However, the effect was attenuated in sensitivity analyses that exclude the region surrounding the _Apo_$\mathcal{E}$ gene, the strongest known risk factor for Alzheimer's disease.[@kim2009] Inclusion of _Apo_$\mathcal{E}$ variants invalidates the exclusion restriction criteria (Assumption 3, above), as the risk reduction observed may be driven by variants in this region via a pathway independent of lipid levels. This was supported by further MR studies in which the observed effect attenuated to the null after _Apo_$\mathcal{E}$ variants were excluded.[@benn2017] Despite the increasing number of MR studies examining this topic, no systematic review of this study design as a source of evidence has been performed.

@@

Despite their advantages, IPD meta-analyses are rarely performed.[@tugwell2010] Factors limiting their uptake include the increased time and effort they require when compared to a summary-level analysis, and the low success rate associated with obtaining the raw data.[@nevitt2017; @ventresca2020] The data underlying primary studies are frequently not publicly available,[@alsheikh-ali2011; @federer2018] and the availability of data "available on request" from authors declines rapidly over time.[@vines2014] Several systematic barriers to open data sharing have been identified[@vanpanhuis2014]. Of particular concern for biomedical IPD analyses are legal issues surrounding the sharing of medical data, motivated by concerns around patient privacy.[@wartenberg2010]

>>>>

[However, evidence from longitudinal studies indicates the challenge of assigning a subtype to dementia outcomes, given the known heterogeneity in clinical phenotypes.]{.correction} [@rooney2017] [For example, ]{.correction} [@shim2013]

@@

Multimorbidity is common among people with dementia, with those diagnosed frequently having two or more other chronic conditions.[@poblador-plou2014]

Patients with dementia have a higher risk of diabetes, stroke, depression pneumonia, falling and infection

[In addition to the patient-centered harm described,]{.correction} dementia also has a substantial economic impact. In 2015, the estimated total cost of dementia in England was £24.2 billion. Health care costs alone were £3.8 billion, with the remainder being divided between unpaid care and social care costs.[@wittenberg2019] Globally, the cost of dementia care is expected to rise to $1tr by 2030.[@prince2014] As such, the urgent need to reduce the burden of dementia, both at the personal and system level, is clear.

@@

Given the substantial burden that dementia represents and the absence of any curative therapies, as detailed in the above sections, the assessment of easily modifiable targets for their utility in the prevention of dementia should be prioritized.[@winblad2016] [Well-studied risk factors for the condition include cardiovascular factors (e.g. hypertension), education, diabetes, and smoking.]{.correction}[@anstey2019] The benefits of a prevention-based approach based on addressing known risk factors for dementia are well-studied. Reducing the prevalence of the seven most important risk factors for dementia (obesity, hypertension, diabetes, smoking, physical inactivity, [depression,]{.correction} and low educational attainment) by 10-20% per decade is estimated to result in a reduction in dementia prevalence of 8-15% by 2050.[@norton2014potential]

In this context [of preventative treatment]{.correction}, lipid levels represent a [potential]{.correction} target, due to the ready availability of lipid-modifying treatments which could be repurposed.[@pushpakom2019] Given the existing uncertainty in the evidence [(as discussed in Section]{.correction} \@ref(evidence-association) [, below),]{.correction} determining whether variations in lipid levels are causative for dementia may prove critical in reducing the future burden of the condition.

@@

Both trials were also limited by the relatively short follow-up period examined. This limited follow-up was to be expected because the primary aim of both trials was to assess the effect of statins on short-term coronary-related outcomes.[@heartprotectionstudycollaborativegroup2002; @trompet2010] The PROSPER trial had a mean follow-up of 3.2 years, while the MRC/BHF [Heart]{.correction} Protection Study estimated risk at 5 years of follow-up. Given the long lag time between non-symptomatic onset of dementia and clinical presentation, it is likely that these durations are insufficient to fully capture the onset of dementia. [Further]{.correction}, as they included only patients at high vascular risk, their generalisability to other settings is limited.[@mcguinness2016]

[A final general limitation of studies of statin use is their limited ability to assess a causal link between lipid levels and dementia. This is caused by the pleiotropic effect of statins, which may effect dementia risk via some non-lipid-lowering pathway. For example, statins influence the levels of C-reactive protein,]{.correction}[@albert2001] [an independent risk factor for dementia.]{.correction}[@lewis2021; @schmidt2002]

@@

Recent MR studies indicated that genetically determined low levels of LDL-c may cause a reduction in AD risk.[@larsson2017; @ostergaard2015] However, the effect was attenuated in sensitivity analyses that exclude the region surrounding the _Apo_$\mathcal{E}$ gene, the strongest known risk factor for Alzheimer's disease.[@kim2009] Inclusion of _Apo_$\mathcal{E}$ variants invalidates the exclusion restriction criteria [of MR analyses], as the risk reduction observed may be driven by variants in this region via a pathway independent of lipid levels. This was supported by further MR studies in which the observed effect attenuated to the null after _Apo_$\mathcal{E}$ variants were excluded.[@benn2017] Despite the increasing number of MR studies examining this topic, no systematic review of this study design as a source of evidence has been performed.

@@

Despite their advantages, IPD meta-analyses are rarely performed.[@tugwell2010] Factors limiting their uptake include the increased time and effort they require when compared to a summary-level analysis, and the low success rate associated with obtaining the raw data.[@nevitt2017; @ventresca2020] The data underlying primary studies are frequently not publicly available,[@alsheikh-ali2011; @federer2018] and the availability of data "available on request" from authors declines rapidly over time.[@vines2014] Several systematic barriers to open data sharing have been identified.[@vanpanhuis2014] Of particular concern for biomedical IPD analyses are legal issues surrounding the sharing of medical data, motivated by concerns around patient privacy.[@wartenberg2010]

### 2022-03-30 03-Chp-Systematic-Review-Methods.Rmd
Range of corrections

One of the limitations of current risk-of-bias assessment practice in systematic reviews is that they are often divorced from the results to which they refer, and are infrequently incorporated into the analysis.[@marusic2020; @katikireddi2015] In response to this criticism, I developed a new visualisation tool which was designed to allow for the production of "paired" forest plots - a risk-of-bias assessment is presented alongside it's corresponding numerical result -  as recommended by the RoB 2 publication.[@sterne2019] In addition, the risk of bias due to missing evidence in each synthesis, as assessed using the RoB-ME tool, is shown beside the overall summary diamond. This tool was developed as an adjunct to this thesis,[@mcguinness2020robvisPaper] and summary of its functionality can be found in Appendix \@ref(appendix-robvis). Unless otherwise stated in the figure, only a single effect measure (hazard ratio/odds ratio/etc.) is represented in a given forest plot.

>>>>

One of the limitations of current risk-of-bias assessment practice in systematic reviews is that they are often divorced from the results to which they refer, and are infrequently incorporated into the analysis.[@marusic2020; @katikireddi2015] In response to this criticism, I developed a new visualisation tool which was designed to allow for the production of "paired" forest plots - [where]{.correction} a risk-of-bias assessment is presented alongside it's corresponding numerical result -  as recommended by the RoB 2 publication.[@sterne2019] In addition, the risk of bias due to missing evidence in each synthesis, as assessed using the RoB-ME tool, is shown beside the overall summary diamond. This tool was developed as an adjunct to this thesis,[@mcguinness2020robvisPaper] and summary of its functionality can be found in Appendix \@ref(appendix-robvis). Unless otherwise stated in the figure, only a single effect measure (hazard ratio/odds ratio/etc.) is represented in a given forest plot.

### 2022-03-30 04-Chp-Systematic-Review-Results.Rmd
Range of corrections

Those records which were excluded in the initial screening but were included by the second reviewer (N =`r discrepancy_Inter`, Table \@ref(tab:agreeInter-table)) were investigated. This discrepancy between the two reviewers was explained in all cases by differing interpretations of the inclusion criteria, most commonly around the definition of cognitive decline versus dementia and of eligible lipids fractions.

@@

The majority of included studies described non-randomised analyses, with the only two included randomised controlled trials (the Heart Protection Study/British Heart Foundation trial[@heartprotectionstudycollaborativegroup2002] and the JUPITER trial[@ridker2008]) both reporting on the effect of statin use on incidence of all-cause dementia in older adults. A similarly small number of Mendelian randomisation studies were identified, several of which employed a two-sample approach using summary statistics from the same published genome wide association studies (GWAS). This leads to complications in the synthesis of these estimates due to the potential for double counting (see Section \@ref(sys-rev-res-AD)).

@@

<!----------------------------------------------------------------------------->
(ref:lipidsDoseResponse-cap) __Dose-response meta-analysis of total cholesterol on Alzheimer's disease__ - Non-randomised studies examining the association of lipids levels with Alzheimer's disease across several categories or "doses" of exposure were synthesised using a dose response meta-analysis. The reference dose was defined _a priori_ as 200 _mg/dL_, the cut-off of the "Normal"/"Optimal" category for total cholesterol as detailed in Table \@ref(tab:lipidLevels-table).

>>>>

Those records which were excluded in the initial screening but were included by the second reviewer (N = `r discrepancy_Inter`, Table \@ref(tab:agreeInter-table)) were investigated. This discrepancy between the two reviewers was explained in all cases by differing interpretations of the inclusion criteria, most commonly around the definition of cognitive decline versus dementia and of eligible lipids fractions.

@@

The majority of included studies described non-randomised [(N = 71) and Mendelian randomisation (N=8)]{.correction} analyses, with the only two included randomised controlled trials (the Heart Protection Study/British Heart Foundation trial[@heartprotectionstudycollaborativegroup2002] and the JUPITER trial[@ridker2008]) both reporting on the effect of statin use on incidence of all-cause dementia in older adults. A similarly small number of Mendelian randomisation studies were identified, several of which employed a two-sample approach using summary statistics from the same published genome wide association studies (GWAS). This leads to complications in the synthesis of these estimates due to the potential for double counting (see Section \@ref(sys-rev-res-AD)).

@@

<!----------------------------------------------------------------------------->
(ref:lipidsDoseResponse-cap) __Dose-response meta-analysis of total cholesterol on Alzheimer's disease__ - Non-randomised studies examining the association of lipids levels with Alzheimer's disease across several categories or "doses" of exposure were synthesised using a dose response meta-analysis. The reference dose was defined _a priori_ as 200 _mg/dL_, the cut-off of the "Normal"/"Optimal" category for total cholesterol as detailed in Table \@ref(tab:lipidLevels-table). [There was little evidence for a non-linear relationship (Wald-type test for deviation from linearity: p = 0.46).]{.correction}

### 2022-03-30 06-Chp-IPD-Analysis.Rmd
Range of corrections

Where data on one of the exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula (Equation \@ref(eq:total-cholesterol-formula)). Lipid levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional statistical power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_.

@@

The results from the main effect analysis across the varying lipid fractions on each dementia outcome considered can be seen in Figures \@ref(fig:mainEffectDem) & \@ref(fig:mainEffectVad), respectively. There was weak evidence for an association of any lipid level with either all-cause or vascular dementia, with the exception of a harmful association between raised triglycerides and vascular dementia (`r ipd_vasdem`, \@ref(fig:mainEffectVad)). For the sole cohort containing data on the Alzheimer's disease outcome (Whitehall II), there was weak evidence for an association of this outcome with any lipid fraction (results shown in comparison with previous analysis of this cohort in Figure \@ref(fig:whitehallComparisonAd))

@@

@@

Investigation revealed several possible reasons for this discrepancy. The first is that while this analysis uses multiple imputation to address missing data, the Tynkkynen _et al._ analysis does not describe how missing data was handled. However, the reduced precision in the reported estimates suggest a complete case analysis. This interpretation is supported by a comparison of summary statistics, which illustrated that my analysis had substantially more dementia events (N =287 in this analysis vs. n=114 in Tynkkynen _et al._), suggesting that incomplete records were discarded in the previous analysis. Finally, the covariates adjusted for in each analysis were similar, though the previous analysis had access to genetic data allowing it to adjust for _Apo_$\mathcal{E}4$ status. However, given that _Apo_$\mathcal{E}4$ is a risk factor for increased LDL-c rather than triglycerides,[@bennet2007] it seems unlikely that additional adjustment for this variable is responsible for the discrepancy in the findings observed.

>>>>

Where data on one of the exposures (TC, LDL-c, HDL-c, TG) was missing, it was inferred from the other three fractions (where available) using the Friedwald formula [as detailed in]{.correction} Equation \@ref(eq:total-cholesterol-formula). Lipid levels were retained as continuous variables rather than dichotomising into a binary hypercholesterolemia exposure, given the additional statistical power this adds to the meta-analysis.[@riley2020;@ensor2018] Where lipid measurements were reported in _mg/dL_, these were converted to _mmol/L_.

@@

The results from the main effect analysis across the varying lipid fractions on each dementia outcome considered can be seen in Figures \@ref(fig:mainEffectDem) & \@ref(fig:mainEffectVad), respectively. There was weak evidence for an association of any lipid level with either all-cause or vascular dementia, with the exception of a harmful association between raised triglycerides and vascular dementia (`r ipd_vasdem`, Figure \@ref(fig:mainEffectVad)). For the sole cohort containing data on the Alzheimer's disease outcome (Whitehall II), there was weak evidence for an association of this outcome with any lipid fraction (results shown in comparison with previous analysis of this cohort in Figure \@ref(fig:whitehallComparisonAd))

@@

#### [Regression dilution bias]{.correction}

[A related issue is regression dilution bias, which occurs when random measurement error in the exposure variable biases the obtained effect estimates towards the null.]{.correction}[@macmahon1990; @hutcheon2010] [In the case of this analysis, the use of a single measurement to define lipid levels may introduce random measurement error into the analysis. As a result, the observed relationship between baseline lipid levels and dementia outcomes could be underestimated. In future research, this bias could be addressed through the averaging of multiple measurements over a given time period to more accurately define the baseline exposure.]{.correction}[@hutcheon2010]

@@

Investigation revealed several possible reasons for this discrepancy. The first is that while this analysis uses multiple imputation to address missing data, the Tynkkynen _et al._ analysis does not describe how missing data was handled. However, the reduced precision in the reported estimates suggest a complete case analysis. This interpretation is supported by a comparison of summary statistics, which illustrated that my analysis had substantially more dementia events (N =287 in this analysis vs. n=114 in Tynkkynen _et al._), suggesting that incomplete records were discarded in the previous analysis. [A further potential cause of the discrepancy is the use of different effect metrics in the two analyses (OR in this analyses, hazard ratios (HR) in Tynkkynen _et al._), though when the outcome is rare, the two measures are expected to be similar (see Section]{.correction} \@ref(sys-rev-analysis-overview) [).]{.correction} Finally, the covariates adjusted for in each analysis were similar, though the previous analysis had access to genetic data allowing it to adjust for _Apo_$\mathcal{E}4$ status. However, given that _Apo_$\mathcal{E}4$ is a risk factor for increased LDL-c rather than triglycerides,[@bennet2007] it seems unlikely that additional adjustment for this variable is responsible for the discrepancy in the findings observed.

### 2022-03-30 07-Chp-Triangulation.Rmd
Range of corrections

<!-- TODO Consider adding AD as an outcome to the IPD row, depending on which result you want to use. -->

\blandscape

<!----------------------------------------------------------------------------->
(ref:thesisOverview-caption) __Summary of research designs included in this thesis__ -The research projects performed as part of this thesis and used as evidence sources in this triangulation exercise are summarised. Note that Chapter \@ref(sys-rev-tools-heading) is intentionally not included in this table, as it describes a research tool rather than a research study.

@@

```{r thesisOverview-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->

\elandscape

>>>>

\blandscape{}
<!----------------------------------------------------------------------------->
(ref:thesisOverview-caption) __Summary of research designs included in this thesis__ -The research projects performed as part of this thesis and used as evidence sources in this triangulation exercise are summarised. Note that Chapter \@ref(sys-rev-tools-heading) is intentionally not included in this table, as it describes a research tool rather than a research study.

@@

```{r thesisOverview-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->
\elandscape{}
